---
title: "Competitive Leslie-Gower in discrete time"
subtitle: ""
author: "W.K. Petry"
date: "`r format(Sys.time(), '%Y-%b-%d')`"
output: github_document
references:
- id: leslie1958biometrika
  title: The properties of a stochastic model for two competing species
  author:
  - family: Leslie
    given: P.H.
  - family: Gower
    given: J.C.
  container-title: Biometrika
  volume: 45
  pages: 316-330
  type: article-journal
  issued:
    year: 1958
- id: cushing2004jdiffeqappl
  title: Some discrete competition models and the competitive exclusion principle
  author:
  - family: Cushing
    given: J.M.
  - family: Leverge
    given: S.
  - family: Chitnis
    given: N.
  - family: Henson
    given: S.M.
  container-title: J Diff Eq Appl
  volume: 10
  pages: 1139-1151
  type: article-journal
  issued:
    year: 2004
nocite: | 
  @cushing2004jdiffeqappl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The model

+--------------------------+---------------------------------------------------------+
| **Number of species**    | >=1                                                     |
+--------------------------+---------------------------------------------------------+
| **Time**                 | discrete                                                |
+--------------------------+---------------------------------------------------------+
| **Interaction type(s)**  | competition                                             |
+--------------------------+---------------------------------------------------------+
| **Difference equation**  | $n_{i,t+1} = \left(\lambda_i n_{i,t}\right)/\left(1+\sum_i\alpha_{ij}n_{j,t}\right)$ |
+--------------------------+---------------------------------------------------------+
| **Parameter meaning**    | $n_{i,t}$: number of individuals of species $i$ at time $t$ |
|                          |                                                         |
|                          | $\lambda_i$: per-capita intrinsic rate of increase of species $i$ |
|                          |                                                         |
|                          | $\alpha_{ij}$: per-capita competitive effect of species $j$ on species $i$ |
+--------------------------+---------------------------------------------------------+
| **Equilibrium**          | $n^*=0$                                                 |
+--------------------------+---------------------------------------------------------+
| **Original publication** | @leslie1958biometrika                                   |
+--------------------------+---------------------------------------------------------+

## Bayesian hierarchical model structure

THe Leslie-Gower model is a multi-species generalization of the Beverton-Holt model.

This framing of the model imagines that you have individual-level data on focal individuals. that includes the number and identity of neighboring competitors ($n_{j,t}$) and the number of offspring produced by the focal individual ($n^{\mathrm{off}}_{i, t}$). The observation of the latent individual fecundities ($f_t$) is a Poisson process. Domain expertise is used to set the priors on $\lambda$ and $\alpha$.

$$
\begin{aligned}
e &= mc^2 \\
c &\sim \mathrm{Normal}\left(299792458, \sigma\right) \\
\end{aligned}
$$

(If this equation looks jumbled, you may need to download the source .Rmd or use a [browser add-on](https://chrome.google.com/webstore/detail/mathjax-plugin-for-github/ioemnmodlmafdkllaclgeombjnmnbima) to render it.)

## Stan code

```{r message=FALSE, warning=FALSE}
## prepare Rmarkdown to input model
library(cmdstanr)  # remotes::install_github("stan-dev/cmdstanr")
register_knitr_engine()
```

```{stan, output.var="temp"}
// Competitive n-species Leslie-Gower in discrete time
// W.K. Petry

// define the structure & types of data to which the model is fit
data {
  int<lower=0> N;                 // number of focal individuals (data rows)
  int<lower=0> nsp;               // number of species
  int<lower=0> focalsp[N];        // species index of the focal individual
  int<lower=0> offspring[N];      // observed offspring production by focal individual
  matrix[N,nsp] bkgd_dens;        // local densities by species (cols) around focal individuals (rows)
  // accept priors from user
  vector[nsp] lambda_mean;        // prior means for lambda
  vector[nsp] lambda_sd;          // prior standard deviations for lambda
  vector[nsp] es_mean;            // prior mean for competitive EFFECTS
  vector[nsp] es_sd;              // prior standard deviation for competitive EFFECTS
  vector[nsp] rs_mean;            // prior mean for competitive RESPONSES
  vector[nsp] rs_sd;              // prior standard deviation for competitive RESPONSES
}

// define the parameters to be fit
parameters {
  vector<lower=0>[nsp] lambda;   // per-capita intrinsic rate of increase
  vector<lower=0>[nsp] es;       // competitive EFFECTS of each species
  vector<lower=0>[nsp] rs;       // competitive RESPONSES of each species
}

transformed parameters {
  matrix[nsp,nsp] alpha = rs * es';  // interaction matrix
  vector<lower=0>[N] fec;        // expected fecundities
  for(i in 1:N) {
    fec[i] = lambda[focalsp[i]] / (1. + bkgd_dens[i] * alpha[focalsp[i]]');
  }
}

model {
  offspring ~ poisson(fec);      // observe expected fecundity
  lambda ~ normal(lambda_mean, lambda_sd);  // normal prior on lambda
  es ~ normal(es_mean, es_sd);   // normal prior on competitive EFFECTS
  rs ~ normal(rs_mean, rs_sd);   // normal prior on competitive RESPONSES
}

```

## Example data & fitting

```{r example, message=FALSE, warning=FALSE}
library(posterior)
library(tidyverse)

# helper function to generate random integer combinations of competitors
rand_vect <- function(N, M, sd = 1, pos.only = TRUE) {
  vec <- rnorm(N, M/N, sd)
  if (abs(sum(vec)) < 0.01) vec <- vec + 1
  vec <- round(vec / sum(vec) * M)
  deviation <- M - sum(vec)
  for (. in seq_len(abs(deviation))) {
    vec[i] <- vec[i <- sample(N, 1)] + sign(deviation)
  }
  if (pos.only) while (any(vec < 0)) {
    negs <- vec < 0
    pos  <- vec > 0
    vec[negs][i] <- vec[negs][i <- sample(sum(negs), 1)] + 1
    vec[pos][i]  <- vec[pos ][i <- sample(sum(pos ), 1)] - 1
  }
  as.integer(vec)
}

## Experimental design
nsp <- 5L  # number of species
bkgd_total <- c(0L, 5L, 50L, 100L, 200L)
reps <- 10L

## Set parameters
set.seed(230342)
lambdas <- rlnorm(n = nsp, meanlog = 7, sdlog = 2) %>%
  `names<-`(letters[1:5])

rs <- rnorm(n = nsp, mean = 1.5, sd = 0.8)  # species per-capita RESPONSES to competitors
es <- rnorm(n = nsp, mean = 1.5, sd = 0.8)  # species per-capita EFFECTS on competitors
alphas <- (rs %o% es) + rnorm(n = nsp^2, mean = 0, sd = 0.25)  # species interaction coefficients
rownames(alphas) <- letters[1:nsp]
colnames(alphas) <- letters[1:nsp]

## Generate data
dat_df <- crossing(sp_focal = letters[1:nsp], bkgd_total, rep = 1:10) %>%
  mutate(bkgd_dens = map(.x = bkgd_total,
                         .f = ~`colnames<-`(t(matrix(rand_vect(nsp, M = .x, sd = .x))),
                                                      paste0("bkgd_", letters[1:nsp]))),
         offspring = map2_int(.x = sp_focal, .y = bkgd_dens,
                              .f = ~rpois(n = 1, as.numeric(lambdas[[.x]] / (1 + alphas[.x, ] %*% t(.y))))),
         bkgd_dens = map(.x = bkgd_dens, .f = data.frame)) %>%  ## help unnest capture column names
  unnest(bkgd_dens)

# prepare data for Stan
dat <- list(N = nrow(dat_df),
            nsp = nsp,
            focalsp = as.integer(factor(dat_df$sp_focal)),
            offspring = dat_df$offspring,
            bkgd_dens = as.matrix(select(dat_df, paste0("bkgd_", letters[1:nsp]))),
            # add priors
            lambda_mean = rep(1000, nsp),
            lambda_sd = rep(200, nsp),
            es_mean = rep(1.5, nsp),
            es_sd = rep(0.8, nsp),
            rs_mean = rep(1.5, nsp),
            rs_sd = rep(0.8, nsp),
            alphasd_scale = 0.25)

## compile the model from an external .stan file
mod <- cmdstan_model("compet_disc_leslie-gower/compet_disc_leslie-gower.stan")

## fit the model (sample the posterior)
ps <- mod$sample(data = dat, refresh = 500, parallel_chains = 2L)
pdat <- as_draws_df(ps$draws(variables = c("alpha", "lambda"))) %>%
  as_tibble()

palphamat <- as_draws_df(ps$draws(variables = c("alpha"))) %>%
  as_tibble() %>%
  pivot_longer(starts_with("alpha"), names_to = c("focal", "bkgd"),
               names_pattern = "^alpha\\[([0-9]+),([0-9]+)\\]", values_to = "alpha",
               names_transform = list(focal = as.integer, bkgd = as.integer)) %>%
  mutate(focal = letters[focal],
         bkgd = letters[bkgd])

plambda <- as_draws_df(ps$draws(variables = c("lambda"))) %>%
  as_tibble() %>%
  pivot_longer(starts_with("lambda"), names_to = "species", names_pattern = "^lambda\\[([0-9]+)\\]",
               values_to = "lambda", names_transform = list(species = as.integer)) %>%
  mutate(species = letters[species])

## examine posterior distribution vs. true values
# c(lambda = lambda, alpha = alpha)  # true values
summarize_draws(ps$draws(variables = c("lambda", "alpha")))  # posterior summary
# 
ggplot(palphamat, aes(x = alpha))+
  geom_density(fill = "cornflowerblue")+
  geom_vline(data = rownames_to_column(as.data.frame(alphas), "focal") %>%
               pivot_longer(-focal, names_to = "bkgd", values_to = "alpha"),
             aes(xintercept = alpha))+
  facet_grid(focal~bkgd, scales = "free_y")+
  theme_minimal(base_size = 20)

ggplot(plambda, aes(x = lambda))+
  geom_density(fill = "cornflowerblue")+
  geom_vline(data = rownames_to_column(as.data.frame(lambdas), "species"),
             aes(xintercept = lambdas))+
  facet_wrap(~species, scale = "free")+
  theme_minimal(base_size = 20)

## Determine pairwise probabilities of competitive outcomes
nstars <- (lambdas - 1) / diag(alphas)

pnstar <- plambda %>%
  full_join(palphamat %>% nest(alphas = c("bkgd", "alpha")),
            by = c(".chain", ".iteration", ".draw", "species" = "focal")) %>%
  mutate(nstar = pmap_dbl(.l = list(..1 = species, ..2 = lambda, ..3 = alphas),
                          .f = ~((..2 - 1) / ..3[..3$bkgd == ..1, "alpha"][[1]]))) %>%
  select(-alphas, -lambda)

pigr <- plambda %>%
  full_join(palphamat %>% nest(alphas = c("bkgd", "alpha")),
            by = c(".chain", ".iteration", ".draw", "species" = "focal")) %>%
  unnest(alphas) %>%
  left_join(pnstar, by = c(".chain", ".iteration", ".draw", "bkgd" = "species")) %>%
  mutate(igr = lambda / (1 + alpha * nstar),
         log_igr = log(igr))

ggplot(pigr %>% filter(species != bkgd), aes(x = log_igr))+
  geom_vline(xintercept = 0, color = "darkgrey")+
  geom_density(fill = "cornflowerblue")+
  facet_grid(species ~ bkgd, scales = "free_y")+
  labs(title = "Posterior distribution of IGR")+
  theme_minimal(base_size = 20)+
  theme(axis.text.y = element_blank())

pc1 <- plambda %>%
  full_join(palphamat %>% nest(alphas = c("bkgd", "alpha")),
            by = c(".chain", ".iteration", ".draw", "species" = "focal")) %>%
  unnest(alphas) %>%
  rename(aij = alpha, lambdai = lambda) %>%
  mutate(ij = paste(species, bkgd, sep = "_"),
         ji = paste(bkgd, species, sep = "_")) %>%
  select(-species, -bkgd)

pchesson <- pc1 %>%
  left_join(pc1 %>%
              select(-ij) %>%
              rename(lambdaj = lambdai, aji = aij, ij = ji)) %>%
  separate(col = "ij", into = c("focal", "bkgd"), sep = "_") %>%
  nest(data = c(lambdai, aij, aji, bkgd, ji, lambdaj)) %>%
  mutate(aii = map2_dbl(.x = data, .y = focal, .f = ~.x[.x$bkgd == .y, "aij"][[1]])) %>%
  unnest(data) %>%
  nest(data = c(lambdai, aii, aij, aji, focal, ji, lambdaj)) %>%
  mutate(ajj = map2_dbl(.x = data, .y = bkgd, .f = ~.x[.x$focal == .y, "aij"][[1]])) %>%
  unnest(data) %>%
  filter(focal != bkgd) %>%
  mutate(fitdiff = ((lambdaj - 1) / (lambdai - 1)) * sqrt((aij * aii) / (ajj * aji)),
         nichediff = 1 - sqrt((aij * aji) / (ajj * aii)))

ggplot(pchesson, aes(x = nichediff, y = fitdiff))+
  geom_point()+
  scale_y_continuous(trans = "pseudo_log")+
  facet_grid(focal ~ bkgd)+
  theme_minimal(base_size = 20)


```

## Variations

-   Change the observation process to allow for over-dispersion

## Session info
```{r}
sessionInfo()
```

## References

::: {#refs}
:::
