---
title: "[Brief, descriptive title]"
subtitle: "([parameterization notes])"
author: "[Your Name Here]"
date: "`r format(Sys.time(), '%Y-%b-%d')`"
output: github_document
references:
- id: extraordinaire2020
  title: An earth-shattering new model for species interactions
  author:
  - family: Extraordinaire
    given: Biomath
  container-title: Journal of Awesome Results
  volume: 1
  type: article-journal
  issued:
    year: 2020
- id: critic2021
  title: Hot takes on the Extraordinaire interaction model
  author:
  - family: Critic
    given: Harshest
  container-title: Journal of Subtweets
  volume: 15
  pages: 1139-1151
  type: article-journal
  issued:
    year: 2021
nocite: | 
  @critic2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The model

[//]: # (FILL in this table with descriptors about the model)

+--------------------------+---------------------------------------------------------+
| **Number of species**    | >=1                                                     |
+--------------------------+---------------------------------------------------------+
| **Time**                 | discrete or continuous                                  |
+--------------------------+---------------------------------------------------------+
| **Interaction type(s)**  | competition                                             |
+--------------------------+---------------------------------------------------------+
| **Difference equation**  | $n_{t+1} = rn_{t}$                                      |
+--------------------------+---------------------------------------------------------+
| **Parameter meaning**    | $n_{t}$: number of individuals at time $t$              |
|                          |                                                         |
|                          | $r$: per-capita intrinsic rate of increase              |
+--------------------------+---------------------------------------------------------+
| **Equilibrium**          | $n^*=0$                                                 |
+--------------------------+---------------------------------------------------------+
| **Original publication** | @extraordinaire2020                                     |
+--------------------------+---------------------------------------------------------+

## Bayesian hierarchical model structure

Describe the hierarchical model here, focusing on the assumed structure & units of the data. In short, what is the high-level view of the experiment or observation study? Be sure to highlight when the data-generating process differs from the classic difference equation in the table above. Please also describe how the observation process is modeled, if relevant. 

$$
\begin{aligned}
e &= mc^2 \\
c &\sim \mathrm{Normal}\left(299792458, \sigma\right) \\
\end{aligned}
$$

(If this equation looks jumbled, you may need to download the source .Rmd or use a [browser add-on](https://chrome.google.com/webstore/detail/mathjax-plugin-for-github/ioemnmodlmafdkllaclgeombjnmnbima) to render it.)

## Stan code

```{r message=FALSE, warning=FALSE}
## prepare Rmarkdown to input model
library(cmdstanr)  # remotes::install_github("stan-dev/cmdstanr")
register_knitr_engine()
```

```{stan, output.var="temp"}
// [MODEL TITLE HERE]
// [YOUR NAME HERE]

// define the structure & types of data to which the model is fit
data {

}

// define the parameters to be fit
parameters {

}

transformed parameters {

}

model {

}

```

## Example data & fitting

```{r example, message=FALSE, warning=FALSE}
library(posterior)
library(ggplot2)

## set 'true' parameter values


## generate synthetic data
set.seed(1234)


dat <- list()

## add priors


## compile the model from an external .stan file
mod <- cmdstan_model("mod.stan")

## fit the model (sample the posterior)
ps <- mod$sample(data = dat, refresh = 500)
pdat <- as_draws_df(ps$draws(variables = c("par1", "par2")))

## examine posterior distribution vs. true values
c(lambda = lambda, alpha = alpha)  # true values
summarize_draws(ps$draws(variables = c("lambda", "alpha")))  # posterior summary

ggplot()+
  geom_hex()+
  geom_point(aes(color = "true value"), shape = 3, stroke = 2, size = 4)+
  scale_color_manual(name = "", values = "black")+
  scale_fill_distiller(name = "posterior density", palette = "RdPu",
                       direction = 1)+
  theme_minimal(base_size = 20)
```

## Variations

-   Suggestions for tuning the model for similar designs

## Session info
```{r}
sessionInfo()
```

## References

::: {#refs}
:::
