---
title: "Consumer-resource Lotka-Volterra in continuous time"
subtitle: ""
author: "W.K. Petry"
date: "`r format(Sys.time(), '%Y-%b-%d')`"
output: github_document
references:
- id: volterra1926
  title: Variazioni e fluttuazioni del numero d’individui in specie animali conviventi
  author:
  - family: Volterra
    given: Vito
  container-title: Memorie Della R. Accademia Dei Lincei, 6th ser.
  volume: 2
  pages: 1-85
  type: article-journal
  issued:
    year: 1926
- id: lotka1925
  title: Principles of physical biology
  author:
  - family: Lotka
    given: A.J.
  publisher: Waverly
  place: Baltimore
  type: book
  issued:
    year: 1925
- id: carpenter2018
  title: 'Predator-Prey Population Dynamics: the Lotka-Volterra model in Stan'
  author:
  - family: Carpenter
    given: Bob
  type: webpage
  url: https://mc-stan.org/users/documentation/case-studies/lotka-volterra-predator-prey.html
  issued:
    year: 2018
- id: mcelreath2020
  title: Statistical rethinking
  author:
  - family: McElreath
    given: Richard
  publisher: CRC Press
  type: book
  edition: 2
  issued:
    year: 2020
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The model

+--------------------------+---------------------------------------------------------+
| **Number of species**    | 2                                                       |
+--------------------------+---------------------------------------------------------+
| **Time**                 | continuous                                              |
+--------------------------+---------------------------------------------------------+
| **Interaction type(s)**  | consumer-resource                                       |
+--------------------------+---------------------------------------------------------+
| **Differential equations** | $\frac{du}{dt}=\alpha u-\beta uv$                     |
|                          |                                                         |
|                          |  $\frac{dv}{dt}=-\gamma v+\delta uv$                    |
+--------------------------+---------------------------------------------------------+
| **Parameter meaning**    | $u$: number of resource/prey                            |
|                          |                                                         |
|                          | $v$: number of consumers/prey                           |
|                          |                                                         |
|                          | $\alpha$: resource/prey growth rate                     |
|                          |                                                         |
|                          | $\beta$: resource loss/predation rate                   |
|                          |                                                         |
|                          | $\delta$: resource/prey conversion rate to consumer growth |
|                          |                                                         |
|                          | $\gamma$: consumer/predator mortality rate              |
+--------------------------+---------------------------------------------------------+
| **Equilibrium**          | $v = \alpha/\beta$                                      |
|                          |                                                         |
|                          | $u = \gamma/\delta$                                     |
+--------------------------+---------------------------------------------------------+
| **Original publications** | @lotka1925 and @volterra1926                           |
+--------------------------+---------------------------------------------------------+

## Bayesian hierarchical model structure

@carpenter2018 provides a thorough introduction to the model and its fit with Stan. Here we simply summarize the key features for completeness, deferring to the original for full detail.

The model assumes that the latent state of the system is the vector $\mathbf{z}=(u,v)$, describing the 'true' abundances of the consumer and resource. The latent state is observed through simultaneous counts the organisms, $\mathbf{y}$, at a particular time.

$$
\begin{aligned}
y_{n,k} &\sim \mathrm{LogNormal}\left(\mathrm{log}(z_{n,k}), \sigma_n\right) \\
\mathbf{z}_n = \left[\begin{matrix}u(t_n)\\v(t_n)\end{matrix}\right] &= \left\{\begin{matrix}\alpha u-\beta uv \\ -\gamma v+\delta uv\end{matrix}\right\} \\
y_{\mathrm{init},k} &\sim \mathrm{LogNormal}\left(\mathrm{log}(z_{\mathrm{init},k}), \sigma_n\right) \\
z_{\mathrm{init, k}} &\sim \mathrm{LogNormal}\left(\mathrm{log}(\mu_{\mathrm{init}}), \sigma_{\mathrm{init}}\right) \\
\alpha, \gamma &\sim \mathrm{Normal}\left(\mu_1, \sigma_1\right) \\
\beta, \delta &\sim \mathrm{Normal}\left(\mu_2, \sigma_2\right) \\
\sigma_n &\sim \mathrm{LogNormal}\left(\mu_{\mathrm{noise}}, \sigma_{\mathrm{noise}}\right)
\end{aligned}
$$

(If this equation looks jumbled, you may need to download the source .Rmd or use a [browser add-on](https://chrome.google.com/webstore/detail/mathjax-plugin-for-github/ioemnmodlmafdkllaclgeombjnmnbima) to render it.)

## Stan code

```{r message=FALSE, warning=FALSE}
## prepare Rmarkdown to input model
library(cmdstanr)  # remotes::install_github("stan-dev/cmdstanr")
register_knitr_engine()
```

```{stan, output.var="temp"}
// Consumer-resource Lotka-Volterra in continuous time
// © 2017–2018, Trustees of Columbia University in New York, licensed under BSD-3.
// originally published: https://mc-stan.org/users/documentation/case-studies/lotka-volterra-predator-prey.html

functions {
  real[] dz_dt(real t,       // time
               real[] z,     // system state {prey, predator}
               real[] theta, // parameters
               real[] x_r,   // unused data
               int[] x_i) {
    real u = z[1];
    real v = z[2];

    real alpha = theta[1];
    real beta = theta[2];
    real gamma = theta[3];
    real delta = theta[4];

    real du_dt = (alpha - beta * v) * u;
    real dv_dt = (-gamma + delta * u) * v;

    return { du_dt, dv_dt };
  }
}
data {
  int<lower = 0> N;          // number of measurement times
  real ts[N];                // measurement times > 0
  real y_init[2];            // initial measured populations
  real<lower = 0> y[N, 2];   // measured populations
}
parameters {
  real<lower = 0> theta[4];   // { alpha, beta, gamma, delta }
  real<lower = 0> z_init[2];  // initial population
  real<lower = 0> sigma[2];   // measurement errors
}
transformed parameters {
  real z[N, 2]
  = integrate_ode_rk45(dz_dt, z_init, 0, ts, theta,
  rep_array(0.0, 0), rep_array(0, 0),
  1e-5, 1e-3, 5e2);
}
model {
  theta[{1, 3}] ~ normal(1, 0.5);
  theta[{2, 4}] ~ normal(0.05, 0.05);
  sigma ~ lognormal(-1, 1);
  z_init ~ lognormal(log(10), 1);
  for (k in 1:2) {
    y_init[k] ~ lognormal(log(z_init[k]), sigma[k]);
    y[ , k] ~ lognormal(log(z[, k]), sigma[k]);
  }
}
generated quantities {
  real y_init_rep[2];
  real y_rep[N, 2];
  for (k in 1:2) {
    y_init_rep[k] = lognormal_rng(log(z_init[k]), sigma[k]);
    for (n in 1:N)
      y_rep[n, k] = lognormal_rng(log(z[n, k]), sigma[k]);
  }
}

```

## Example data & fitting

We will fit the model to real-world data of lynx and hare pelts trapped by the Hudson's Bay Company between 1900-1920 (*note: these data have gone through a series of telephone game-like cycles of being extracted from figures. Caveat emptor.*) The code below has been modified from @carpenter2018.

```{r example, message=FALSE, warning=FALSE}
library(posterior)
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)

## read in data
lynx_hare_df <- read.csv("https://raw.githubusercontent.com/stan-dev/example-models/master/knitr/lotka-volterra/hudson-bay-lynx-hare.csv",
                         comment.char="#")

## prepare data for Stan
N <- length(lynx_hare_df$Year) - 1
ts <- 1:N
y_init <- c(lynx_hare_df$Hare[1], lynx_hare_df$Lynx[1])
y <- as.matrix(lynx_hare_df[2:(N + 1), 2:3])
y <- cbind(y[ , 2], y[ , 1]); # hare, lynx order
dat <- list(N = N, ts = ts, y_init = y_init, y = y)

## compile the model from an external .stan file
mod <- cmdstan_model("conres_cont_lotka-volterra.stan")

## fit the model (sample the posterior)
ps <- mod$sample(data = dat, refresh = 500, seed = 23003)
pdraws <- rename_variables(ps$draws(), alpha = "theta[1]", beta = "theta[2]", gamma = "theta[3]",
                         delta = "theta[4]", `z[0,1]` = "z_init[1]", `z[0,2]` = "z_init[2]")
pdat <- as_draws_df(pdraws)

pts <- pdat %>%
  as_tibble() %>%
  select(.draw, starts_with("z[")) %>%
  pivot_longer(cols = -.draw, names_to = c("time", "species"),
               names_pattern = "z\\[([0-9]+),([1-2])\\]", values_to = "latent_N",
               names_transform = list(time = as.integer)) %>%
  mutate(time = 1900 + time,
         species = case_when(
           species == "1" ~ "hare",
           species == "2" ~ "lynx"
         ))

## examine pairwise views of posterior distribution
summarize_draws(pdraws)  # posterior summary

## examine the posterior latent timeseries (+ observed pelt counts)
ggplot(pts, aes(x = time, y = latent_N, group = paste0(.draw, species), color = species))+
  geom_line(alpha = 0.05)+
  geom_line(data = lynx_hare_df, aes(x = Year, y = Hare, color = "pelts hare"), inherit.aes = FALSE,
            size = 1.75)+
  geom_line(data = lynx_hare_df, aes(x = Year, y = Lynx, color = "pelts lynx"), inherit.aes = FALSE,
            size = 1.75)+
  labs(title = "Posterior time series", y = "N")+
  scale_color_manual(values = c("#37b792", "#379cb7", "#257b62", "#25697b"),
                     guide = guide_legend(override.aes = list(size = c(1, 1, 2, 2),
                                                              alpha = c(0.5, 0.5, 1, 1))))+
  theme_minimal(base_size = 20)

## determine the period of the oscillations around the fixed point
period <- (2 * pi) / sqrt(pdat$alpha * pdat$gamma)

ggplot(data = NULL, aes(x = period))+
  geom_density(fill = "cornflowerblue")+
  labs(title = "Posterior of the oscillation period", x = "Period (years)")+
  theme_minimal(base_size = 20)+
  theme(axis.text.y = element_blank())
```

## Variations

-   Add a lag variable, following @mcelreath2020 (p. 551)
-   Add observation (trap) probability parameter to prevent the latent number from being less than the observed number of pelts, following @mcelreath2020 (p. 545-550)

## Session info
```{r}
sessionInfo()
```

## References

::: {#refs}
:::
