---
title: "Competitive Beverton-Holt"
subtitle: "(competition coefficient parameterization)"
author: "W.K. Petry"
date: "`r format(Sys.time(), '%Y-%b-%d')`"
output: rmarkdown::github_document
references:
- id: beverton1957
  title: On the Dynamics of Exploited Fish Populations
  author:
  - family: Beverton
    given: R.J.
  - family: Holt
    given: S.J.
  container-title: Fishery Investigations Series II
  volume: XIX
  publisher: Ministry of Agriculture, Fisheries and Food
  type: book
  issued:
    year: 1957
- id: cushing2004jdiffeqappl
  title: Some discrete competition models and the competitive exclusion principle
  author:
  - family: Cushing
    given: J.M.
  - family: Leverge
    given: S.
  - family: Chitnis
    given: N.
  - family: Henson
    given: S.M.
  container-title: J Diff Eq Appl
  volume: 10
  pages: 1139-1151
  type: article-journal
  issued:
    year: 2004
nocite: | 
  @cushing2004jdiffeqappl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The model

+--------------------------+---------------------------------------------------------+
| **Number of species**    | 1                                                       |
+--------------------------+---------------------------------------------------------+
| **Time**                 | discrete                                                |
+--------------------------+---------------------------------------------------------+
| **Interaction type(s)**  | competition                                             |
+--------------------------+---------------------------------------------------------+
| **Difference equation**  | $n_{t+1} = n_{t}\lambda / (1+\alpha n_{t})$             |
+--------------------------+---------------------------------------------------------+
| **Parameter meaning**    | $n_{t}$: number of individuals at time $t$              |
|                          |                                                         |
|                          | $\lambda$: per-capita intrinsic rate of increase        |
|                          |                                                         |
|                          | $\alpha$: (intra-specific) competition coefficient      |
+--------------------------+---------------------------------------------------------+
| **Equilibrium**          | $n^*=(\lambda-1)/\alpha$                                |
+--------------------------+---------------------------------------------------------+
| **Original publication** | @beverton1957, using carrying capacity parameterization |
+--------------------------+---------------------------------------------------------+

## Bayesian hierarchical model structure

This framing of the model imagines that you have individual-level data on focal individuals. that includes the number of neighboring competitors ($n_t$) and the number of offspring produced ($n_{\mathrm{off}, t}$). The observation of the latent individual fecundities ($f_t$) is a Poisson process. Domain expertise is used to set the priors on $\lambda$ and $\alpha$.

$$
\begin{aligned}
n_{\mathrm{off}, t} &\sim \mathrm{Poisson}\left(f_{t}\right) \\
f_t &= \frac{\lambda}{1+\alpha n_{t}} \\
\lambda &\sim \mathrm{Normal}\left(\mu_{\lambda},\sigma_{\lambda}\right) \\
\alpha &\sim \mathrm{Normal}\left(\mu_{\alpha},\sigma_{\alpha}\right).
\end{aligned}
$$

## Stan code

```{r message=FALSE, warning=FALSE}
## prepare Rmarkdown to input model
library(cmdstanr)  # remotes::install_github("stan-dev/cmdstanr")
register_knitr_engine()
```

```{stan, output.var="temp"}
// Beverton-Holt competition model
// W.K. Petry

// define the structure & types of data to which the model is fit
data {
  int<lower=0> N;                 // number of focal individuals (data rows)
  int<lower=0> offspring[N];      // observed offspring production by focals
  int<lower=0> bkgd_dens[N];      // local densities around focal individuals
  // accept priors from user
  real<lower=0> lambda_mean;      // prior mean for lambda
  real<lower=0> lambda_sd;        // prior standard deviation for lambda
  real<lower=0> alpha_mean;       // prior mean for competition coefficient
  real<lower=0> alpha_sd;         // prior standard deviation for competition coefficent
}

// define the parameters to be fit
parameters {
  real<lower=0> lambda;          // per-capita intrinsic rate of increase
  real<lower=0> alpha;           // competition coefficient
}

transformed parameters {
  vector[N] fec;                 // turn data-generating process crank
  for (i in 1:N) {               // to get latent fecundity
    fec[i] = lambda / (1. + alpha * bkgd_dens[i]);
  }
}

model {
  offspring ~ poisson(fec);     // observe latent fecundity as whole offspring
  lambda ~ normal(lambda_mean, lambda_sd);  // normal prior on lambda
  alpha ~ normal(alpha_mean, alpha_sd);     // normal prior on alpha
}

```

## Example data & fitting

```{r example, message=FALSE, warning=FALSE}
library(posterior)
library(ggplot2)

## set 'true' parameter values
lambda <- 200
alpha <- 1.25

## generate synthetic data
set.seed(58428)
N <- 500L  # number of observed individuals
bkgd_dens <- as.integer(round(runif(n = N, min = 0L, max = 100)))
offspring <- rpois(n = N, lambda = (lambda / (1 + alpha * bkgd_dens)))

dat <- list(N = N, bkgd_dens = bkgd_dens, offspring = offspring)

## add priors
## (note both parameters are truncated at zero by the Stan code)
dat$lambda_mean <- 100
dat$lambda_sd <- 50
dat$alpha_mean <- 0
dat$alpha_sd <- 1

## compile the model from an external .stan file
bh <- cmdstan_model("bh.stan")

## fit the model (sample the posterior)
ps <- bh$sample(data = dat, refresh = 500)
pdat <- as_draws_df(ps$draws(variables = c("lambda", "alpha")))

## plot posterior
ggplot(data = NULL, aes(x = alpha, y = lambda))+
  geom_hex(data = pdat)+
  geom_point(aes(color = "true value"), shape = 3, stroke = 2, size = 4)+
  scale_color_manual(name = "", values = "black")+
  scale_fill_distiller(name = "posterior density", palette = "RdPu",
                       direction = 1)+
  theme_minimal(base_size = 20)
```

## Variations

-   Change the observation process to allow for overdispersion (e.g., negative binomial)

-   Set heavier-tailed prior on $\alpha$ (e.g., Student's t)

-   Set more flexible prior on $\lambda$ (e.g., log-normal)

## Session info
```{r}
sessionInfo()
```

## References

::: {#refs}
:::
